Exploratory Analysis Script

### 2024-03-22 code/eda-code/eda-v1.qmd
Debugged Code Part 2
Pushing it to get some feedback

```{r, message=FALSE,  warning=FALSE}
#load needed packages. make sure they are installed.
library(tidyverse) #for data processing/cleaning; includes ggplot2, tidyr, readr, dplyr, stringr, purr, forcats
library(skimr) #for nice visualization of data 
library(here) #to set paths
library(renv) #for package management
library(knitr) #for nice tables
library(kableExtra) #for nice tables
library(gt) #for nice tables
library(gtsummary) #for summary tables
library(ggplot2) #for plotting
library(hrbrthemes) #for nice themes
hrbrthemes::import_roboto_condensed() #import needed fonts for hrbrthemes
#rm(list=ls()) #clear the environment
```

@@

```{r}
#Path to summary data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","ML_summary_processed.rds")
#load data
ML_summary <- readRDS(data_location)
```

@@

Before we create a summary table, lets clean up our variables for easier analysis. First, we need to create dummy variables for categories. We will start with the `Color` variable, which is maternal skin color. We will also create dummy variables for the `Status` and `Nationality` variables for ease of interpretation. The original data categorized maternal racial categories as White (*Branca*), the reference group here; mixed-race (*Parda*); and Black (*Preta*). I maintained this categorization. However, I also created a new dummy variable for skin color, combining Black and Mixed-Race patients into one category, `Afro-Descent` and all White patients into a `Euro-Descent` category. For the `Status` variable, we will combine "multigesta", "secundigesta", and "trigesta" into "multigravida", since they mean the same thing (two or more pregnancies) and change the language to English (from gesta to gravida). We will also combine `secundipara` and `multipara` into one category `multipara`, since this means two or more viable pregnancies. The words in Portuguese and English here are the same. For the `Nationality` variable, we will combine all European nationalities into one category, "European", all Latin American nationalities into "LatinAmerican", and all Middle Eastern nationalities into "MiddleEastern". I am combining categories because for many of the `European` and `Latin American` categories, there are very few observations. Moreover, I am interested in immigration as a general status rather than the relationship between specific immigrant countries and health status in Brazil. "Brazilian" will remain the same.

@@

```{r}
# Create dummy variables
# Create dummy variables using recode factor from dplyr
 ML_linear$ModifiedColor <- recode_factor(ML_linear$Color, "Parda" = "AfroDescent", "Preta" = "AfroDescent", "Branca" = "EuroDescent")

@@

# View the updated dataframe
print(head(ML_linear))

# Change observations from Portuguese to English
ML_linear <- ML_linear %>%
  mutate(
    ModifiedColor = recode_factor(ModifiedColor, "AfroDescent" = "Afro-Descent", "EuroDescent" = "Euro-Descent"),
    ModifiedNationality = recode_factor(ModifiedNationality, "European" = "European", "LatinAmerican" = "Latin American", "MiddleEastern" = "Middle Eastern", "Brazilian" = "Brazilian"),
    Color = recode_factor(Color, "Preta" = "Black", "Parda" = "Mixed Race", "Branca" = "White"),
    Nationality = recode_factor(Nationality, "Alema" = "German", "Argentina" = "Argentine", "Austriaca" = "Austrian", "Brasileira" = "Brazilian", "Espanhola" = "Spanish", "Francesa" = "French", "Italiana" = "Italian", "Paraguaya" = "Paraguayan", "Polaca" = "Polish", "Portuguesa" = "Portuguese", "Rumania" = "Romanian", "Russa" = "Russian", "Suiça" = "Swiss", "Siria" = "Syrian", "Uruguaya" = "Uruguayan"),
    Birth = recode_factor(Birth, "aborto" = "Abortion", "intervencionista" = "Interventionist", "natural" = "Natural", "operatório" = "Operative"),
    MaternalOutcome = recode_factor(MaternalOutcome, "alta" = "Discharged", "morta" = "Death", "transferência" = "Hospital transferal"),
    FetalOutcome = recode_factor(FetalOutcome, "vivo" = "Live Birth", "morto" = "Stillbirth or Neonatal Death")
    )
```

>>>>

```{r, message=FALSE,  warning=FALSE}
#load needed packages. make sure they are installed.
library(tidyverse) #for data processing/cleaning; includes ggplot2, tidyr, readr, dplyr, stringr, purr, forcats
library(skimr) #for nice visualization of data 
library(here) #to set paths
library(renv) #for package management
library(knitr) #for nice tables
library(kableExtra) #for nice tables
library(gt) #for nice tables
library(gtsummary) #for summary tables
library(ggplot2) #for plotting
library(hrbrthemes) #for nice themes
hrbrthemes::import_roboto_condensed() #import needed fonts for hrbrthemes
rm(list=ls()) #clear the environment
```

@@

```{r}
#Path to summary data. Note the use of the here() package and not absolute paths
data_location <- here::here("data","processed-data","ML_summary_processed.rds")

#load data
ML_summary <- readRDS(data_location)
```

@@

Before we create a summary table, lets clean up our variables for easier analysis. First, we need to create dummy variables for categories. We will start with the `Color` variable, which is maternal skin color. We will also create dummy variables for the `Status` and `Nationality` variables for ease of interpretation. The original data categorized maternal racial categories as White (*branca*), the reference group here; mixed-race (*parda*); and Black (*preta*). I maintained this categorization. However, I also created a new dummy variable for skin color, combining Black and Mixed-Race patients into one category, `Afro-Descent` and all White patients into a `Euro-Descent` category. For the `Status` variable, we will combine "multigesta", "secundigesta", and "trigesta" into "multigravida", since they mean the same thing (two or more pregnancies) and change the language to English (from gesta to gravida). We will also combine `secundipara` and `multipara` into one category `multipara`, since this means two or more viable pregnancies. The words in Portuguese and English here are the same. For the `Nationality` variable, we will combine all European nationalities into one category, "European", all Latin American nationalities into "LatinAmerican", and all Middle Eastern nationalities into "MiddleEastern". I am combining categories because for many of the `European` and `Latin American` categories, there are very few observations. Moreover, I am interested in immigration as a general status rather than the relationship between specific immigrant countries and health status in Brazil. "Brazilian" will remain the same.

@@

#Check pre-structure
str(ML_linear)

# Create dummy variables using recode factor from dplyr
 ML_linear$ModifiedColor <- recode_factor(ML_linear$Color, "Parda" = "AfroDescent", "Preta" = "AfroDescent", "Branca" = "EuroDescent")

@@

# Change observations from Portuguese to English
ML_linear <- ML_linear %>%
  mutate(
    ModifiedColor = recode_factor(ModifiedColor, "AfroDescent" = "Afro-Descent", "EuroDescent" = "Euro-Descent"),
    ModifiedNationality = recode_factor(ModifiedNationality, "European" = "European", "LatinAmerican" = "Latin American", "MiddleEastern" = "Middle Eastern", "Brazilian" = "Brazilian"),
    Color = recode_factor(Color, "Preta" = "Black", "Parda" = "Mixed Race", "Branca" = "White"),
    Nationality = recode_factor(Nationality, "Alema" = "German", "Argentina" = "Argentine", "Austriaca" = "Austrian", "Brasileira" = "Brazilian", "Espanhola" = "Spanish", "Francesa" = "French", "Italiana" = "Italian", "Paraguaya" = "Paraguayan", "Polaca" = "Polish", "Portuguesa" = "Portuguese", "Rumania" = "Romanian", "Russa" = "Russian", "Suiça" = "Swiss", "Siria" = "Syrian", "Uruguaya" = "Uruguayan"),
    Birth = recode_factor(Birth, "aborto" = "Abortion", "intervencionista" = "Interventionist", "natural" = "Natural", "operatório" = "Operative"),
    MaternalOutcome = recode_factor(MaternalOutcome, "alta" = "Discharged", "morta" = "Death", "transferência" = "Hospital transferal"),
    FetalOutcome = recode_factor(FetalOutcome, "vivo" = "Live Birth", "morto" = "Stillbirth or Neonatal Death")
    )
str(ML_linear)
```
